---
title: "Homework 13"
author: "张子乐, 3230104237"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage[fontset=fandol]{ctex}
  - \usepackage{float}
---

```{r setup, include=FALSE}
Sys.setenv(FONTCONFIG_PATH = "C:/Users/admin/AppData/Roaming/TinyTeX/tlpkg/tlpostcode/xetex/conf")
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  dev = "cairo_pdf",
  comment = "",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 5.5,
  fig.height = 3.4,
  out.width = "75%"
)

library(ggplot2)
library(survival)
library(flexsurv)
```

# Practical 12: Survival Analysis

## Answer to Problem 1

### (a) *Is the Censoring Mechanism Uninformative?*

Let $f_C$ and $S_C$ denote the density function and survivor function of the censoring time $C_i$.

- If $\delta_i=1$, then the event is observed at time $y_i$, meaning that $T_i=y_i, C_i\ge y_i.$ Since $T_i$ and $C_i$ are independent, the joint contribution is $f_T(y_i;\lambda)S_C(y_i).$

- If $\delta_i=0$, then censoring occurs at time $y_i$, meaning that $C_i=y_i,T_i>y_i.$ The joint contribution is $f_C(y_i)S_T(y_i;\lambda).$

Thus, the likelihood is

$$
L(\lambda,\eta)
=
\prod_{i=1}^n
\left[
f_T(y_i;\lambda)S_C(y_i;\eta)
\right]^{\delta_i}
\left[
S_T(y_i;\lambda)f_C(y_i;\eta)
\right]^{1-\delta_i},
$$

where $\eta$ denotes the parameters of the censoring distribution.

This can be factorized as

$$
L(\lambda,\eta)
=
\left\{
\prod_{i=1}^n
f_T(y_i;\lambda)^{\delta_i}
S_T(y_i;\lambda)^{1-\delta_i}
\right\}
\left\{
\prod_{i=1}^n
S_C(y_i;\eta)^{\delta_i}
f_C(y_i;\eta)^{1-\delta_i}
\right\}.
$$

The first factor depends only on $\lambda$, while the second factor depends only on the censoring parameters $\eta$.

According to the condition given in the problem:

*$C_i$ is independent of $T_i$, and that the parameters of the distribution of $C_i$ are distinct from the parameter $\lambda$*

Therefore, the censoring mechanism is **uninformative** for inference about $\lambda$.

* * * 

### (b) *MLE of $\lambda$ and Its Asymptotic Distribution*

Since $T_i\sim\operatorname{Exp}(\lambda)$, we have

- $h(t)=\lambda$ and $H(t)=\lambda t$
- $S_T(t)=e^{-\lambda t}$ and $f_T(t)=\lambda e^{-\lambda t}$

The likelihood function is
$$
\begin{aligned}
L(\lambda)
&=
\prod_{i=1}^n
f_T(y_i;\lambda)^{\delta_i}
S_T(y_i;\lambda)^{1-\delta_i}\\
&=
\prod_{i=1}^n
\left[\lambda e^{-\lambda y_i}\right]^{\delta_i}
\left[e^{-\lambda y_i}\right]^{1-\delta_i}\\
&=
\prod_{i=1}^n
\lambda^{\delta_i}e^{-\lambda y_i}\\
&=
\lambda^{\sum_{i=1}^n\delta_i}
\exp\left(
-\lambda\sum_{i=1}^n y_i
\right)\\
&=
\lambda^D e^{-\lambda R}
\end{aligned}
$$
where $D=\sum_{i=1}^n\delta_i$ and $R=\sum_{i=1}^n y_i.$

The log-likelihood function is
$$
\ell(\lambda)
=
D\log\lambda-\lambda R.
$$

Differentiating with respect to $\lambda$ gives

$$
\frac{\partial\ell(\lambda)}{\partial\lambda}
=
\frac{D}{\lambda}-R.
$$

Then we obtain

$$
\boxed{
\widehat{\lambda}_{\text{MLE}}
=
\frac{D}{R}
=
\frac{\sum_{i=1}^n\delta_i}
{\sum_{i=1}^n y_i}.
}
$$

If $D=0$, the likelihood is maximized in the limiting sense as $\widehat{\lambda}\to 0$.

As for the asymptotic distribution, we use the corollary of MLE.

For one observation, the log-likelihood is

$$
\ell_i(\lambda)
=
\delta_i\log\lambda-\lambda y_i.
$$

The second derivative is

$$
\frac{\partial^2\ell_i(\lambda)}{\partial\lambda^2}
=
-\frac{\delta_i}{\lambda^2}.
$$

Therefore, the Fisher information from one observation is

$$
\begin{aligned}
\mathcal I_1(\lambda)
&=
-
E\left[
\frac{\partial^2\ell_i(\lambda)}{\partial\lambda^2}
\right]\\
&=
\frac{E(\Delta_i)}{\lambda^2}.
\end{aligned}
$$

Define

$$
p=P(\Delta_i=1)=P(T_i\le C_i).
$$

Then

$$
\boxed{
\mathcal I_1(\lambda)
=
\frac{p}{\lambda^2}.
}
$$

Since $T_i$ and $C_i$ are independent,

$$
\begin{aligned}
p
&=
P(T_i\le C_i)\\
&=
\int_0^\infty
f_T(t)P(C_i\ge t)\,dt\\
&=
\int_0^\infty
\lambda e^{-\lambda t}S_C(t)\,dt.
\end{aligned}
$$



By the asymptotic normality of the MLE,

$$
\sqrt{n}(\widehat{\lambda}-\lambda)
\xrightarrow{d}
N\left(
0,\,
\mathcal I_1(\lambda)^{-1}
\right).
$$

we obtain

$$
\boxed{
\sqrt{n}(\widehat{\lambda}-\lambda)
\xrightarrow{d}
N\left(
0,\,
\frac{\lambda^2}{p}
\right).
}
$$

Equivalently,

$$
\boxed{
\widehat{\lambda}
\overset{a}{\sim}
N\left(
\lambda,\,
\frac{\lambda^2}{np}
\right).
}
$$


Here $p=P(\Delta_i=1)$ is the probability that the event is observed. 
Since the censoring distribution is not specified, $p$ can not be reduced to a numerical value, 
and we estimate it by $\widehat{p}=\frac{D}{n}$, where $D=\sum_{i=1}^n\delta_i$.

An estimated variance is

$$
\widehat{\operatorname{Var}}(\widehat{\lambda})
\approx
\frac{\widehat{\lambda}^2}{D},
$$

When there is no censoring, $p=1$, and the result reduces to

$$
\widehat{\lambda}
\overset{a}{\sim}
N\left(
\lambda,\,
\frac{\lambda^2}{n}
\right).
$$

* * *

### (c) *the PDF and Survivor Function for $T_i$*

It's easy to find that
$$
\boxed{
H(t)
=
\begin{cases}
\lambda_1 t, & 0\le t<\tau_1,\\[4pt]
\lambda_1\tau_1+\lambda_2(t-\tau_1), & t\ge \tau_1.
\end{cases}
}
$$

Since

$$
S_T(t)=\exp\{-H(t)\},
$$

we have

$$
\boxed{
S_T(t)
=
\begin{cases}
\exp(-\lambda_1t), & 0\le t<\tau_1,\\[6pt]
\exp\left[-\lambda_1\tau_1-\lambda_2(t-\tau_1)\right],
& t\ge\tau_1.
\end{cases}
}
$$

Since $f_T(t)=h(t)S_T(t),$ we obtain
$$
\boxed{
f_T(t)
=
\begin{cases}
\lambda_1\exp(-\lambda_1t),
& 0\le t<\tau_1,\\[6pt]
\lambda_2
\exp\left[-\lambda_1\tau_1-\lambda_2(t-\tau_1)\right],
& t\ge\tau_1.
\end{cases}
}
$$

Equivalently, in terms of $(\nu_1,\nu_2)$,

$$
\boxed{
f_T(t)
=
\begin{cases}
e^{\nu_1}\exp(-e^{\nu_1}t),
& 0\le t<\tau_1,\\[6pt]
e^{\nu_2}
\exp\left[-e^{\nu_1}\tau_1-e^{\nu_2}(t-\tau_1)\right],
& t\ge\tau_1.
\end{cases}
}
$$

---

### (d) *MLE of $(\nu_1,\nu_2)$ and Its Asymptotic Distribution*

Define

$$
I_{ij}
=
\begin{cases}
1, & \tau_{j-1}\le y_i<\tau_j,\\
0, & \text{otherwise},
\end{cases}
\qquad j=1,2.
$$

which indicates whether the observed time $y_i$ falls in interval $j$.

For the two intervals,

$$
I_{i1}=I(0\le y_i<\tau_1),
\qquad
I_{i2}=I(y_i\ge \tau_1).
$$

Define the time spent by individual $i$ in interval $j$ as

$$
y_{ij}
=
\max\left\{
0,\,
\min\left(\tau_j-\tau_{j-1},\,y_i-\tau_{j-1}\right)
\right\}.
$$

In the present two-interval case,

$$
y_{i1}=\min(y_i,\tau_1)
$$

and

$$
y_{i2}=(y_i-\tau_1)_+
=
\max(0,y_i-\tau_1).
$$

Now define

$$
d_j=\sum_{i=1}^n \delta_i I_{ij},
\qquad
r_j=\sum_{i=1}^n y_{ij},
\qquad j=1,2.
$$

Here,

- $d_j$ is the number of observed events in interval $j$
- $r_j$ is the total time at risk contributed in interval $j$

For individual $i$, the likelihood related to the survival parameters is

$$
L_i(\nu_1,\nu_2)
=
h(y_i)^{\delta_i}\exp\{-H(y_i)\}.
$$

Using the interval notation, this can be written as

$$
L_i(\nu_1,\nu_2)
=
\prod_{j=1}^2
\left(e^{\nu_j}\right)^{\delta_i I_{ij}}
\exp\left(-e^{\nu_j}y_{ij}\right).
$$

Therefore, the likelihood function is

$$
\begin{aligned}
L(\nu_1,\nu_2)
&=
\prod_{i=1}^n
\prod_{j=1}^2
\left(e^{\nu_j}\right)^{\delta_i I_{ij}}
\exp\left(-e^{\nu_j}y_{ij}\right)\\
&=
\prod_{j=1}^2
\left(e^{\nu_j}\right)^{d_j}
\exp\left(-e^{\nu_j}r_j\right).
\end{aligned}
$$

Hence, the log-likelihood function is

$$
\ell(\nu_1,\nu_2)
=
d_1\nu_1-r_1e^{\nu_1}
+
d_2\nu_2-r_2e^{\nu_2}.
$$

For $j=1,2$,

$$
\frac{\partial\ell}{\partial\nu_j}
=
d_j-r_je^{\nu_j}.
$$

Therefore,

$$
e^{\widehat{\nu}_j}
=
\frac{d_j}{r_j}.
$$

Since $\lambda_j=e^{\nu_j}$, we obtain the MLEs

$$
\boxed{
\widehat{\lambda}_j
=
\frac{d_j}{r_j},
\widehat{\nu}_j
=
\log\left(\frac{d_j}{r_j}\right),
j=1,2.
}
$$

That is,

$$
\boxed{
\widehat{\nu}_1
=
\log\left(
\frac{\sum_{i=1}^n\delta_i I_{i1}}
{\sum_{i=1}^n y_{i1}}
\right), 
\widehat{\nu}_2
=
\log\left(
\frac{\sum_{i=1}^n\delta_i I_{i2}}
{\sum_{i=1}^n y_{i2}}
\right).
}
$$

If $d_j=0$ for some interval $j$, then $\widehat{\lambda}_j=0$
and $\widehat{\nu}_j=-\infty.$
Thus, there is no finite interior MLE for $\nu_j$ when no observed event occurs in interval $j$.

The second derivatives of the log-likelihood are

$$
\frac{\partial^2\ell}{\partial\nu_j^2}
=
-r_je^{\nu_j},
\qquad j=1,2,
$$

and

$$
\frac{\partial^2\ell}{\partial\nu_1\partial\nu_2}
=
0.
$$

For one observation, define

$$
m_j=E(Y_{ij}),
\qquad j=1,2.
$$

Then the Fisher information contribution for $\nu_j$ is

$$
\mathcal I_j
=
e^{\nu_j}m_j.
$$

Define

$$
\pi_j
=
P\left(
\Delta_i=1,\,
\tau_{j-1}\le Y_i<\tau_j
\right),
\qquad j=1,2.
$$

This is the probability that an individual has an observed event in interval $j$.

Under independent censoring, $\pi_j=e^{\nu_j}E(Y_{ij}),$ the Fisher information matrix for one observation is

$$
\mathcal I(\nu_1,\nu_2)
=
\begin{pmatrix}
\pi_1 & 0\\
0 & \pi_2
\end{pmatrix}
$$

Where

$$
\pi_1
=
\int_0^{\tau_1}
e^{\nu_1}
\exp(-e^{\nu_1}t)
S_C(t)\,dt,
\pi_2
=
\int_{\tau_1}^{\infty}
e^{\nu_2}
\exp\left[
-e^{\nu_1}\tau_1-e^{\nu_2}(t-\tau_1)
\right]
S_C(t)\,dt.
$$

Assume that $\pi_1>0$ and $\pi_2>0.$

Then, by the asymptotic normality of the MLE,

$$
\boxed{
\sqrt{n}
\begin{pmatrix}
\widehat{\nu}_1-\nu_1\\
\widehat{\nu}_2-\nu_2
\end{pmatrix}
\xrightarrow{d}
N_2
\left[
\begin{pmatrix}
0\\
0
\end{pmatrix},
\begin{pmatrix}
1/\pi_1 & 0\\
0 & 1/\pi_2
\end{pmatrix}
\right].
}
$$

Equivalently, for each $j=1,2$,

$$
\boxed{
\widehat{\nu}_j
\overset{a}{\sim}
N\left(
\nu_j,\,
\frac{1}{n\pi_j}
\right).
}
$$

Since $\widehat{\pi}_j=\frac{d_j}{n},$ we may estimate the asymptotic variance by $\widehat{\operatorname{Var}}(\widehat{\nu}_j)\approx\frac{1}{d_j}.$

As for *why use the parameterization $\lambda_j=e^{\nu_j}$*,
the hazard rates must satisfy $\lambda_j>0$.
If we estimate $\lambda_j$ directly, the optimization problem must be constrained to positive values.

By writing $\lambda_j=e^{\nu_j},$
we allow $\nu_j\in\mathbb R$ 
without any additional constraints, while automatically ensuring that $\lambda_j>0.$

This parameterization also gives a convenient log-likelihood:

$$
\ell(\nu_1,\nu_2)
=
\sum_{j=1}^2
\left[
d_j\nu_j-r_je^{\nu_j}
\right],
$$

which is concave in $(\nu_1,\nu_2)$.

---

### Additional question: *Is $T_i$ being independent of $C_i$ sufficient for censoring to be uninformative?*

(*Consider the situation where $T_i$ is independent of $C_i$, but both $T_i$ and $C_i$ are distributed as Exponential random variables with common rate parameter $\lambda$.*)

**NOT SUFFICIENT.**

Under independent censoring, the full likelihood can be factorized as

$$
L(\lambda,\eta)
=
\left\{
\prod_{i=1}^n
f_T(y_i;\lambda)^{\delta_i}
S_T(y_i;\lambda)^{1-\delta_i}
\right\}
\left\{
\prod_{i=1}^n
S_C(y_i;\eta)^{\delta_i}
f_C(y_i;\eta)^{1-\delta_i}
\right\}.
$$

If the censoring parameter $\eta$ is distinct from $\lambda$, then the second factor does not involve $\lambda$. 
Therefore, when estimating $\lambda$, it can be ignored. In this case, the censoring mechanism is uninformative.

However, suppose that $T_i\sim \operatorname{Exp}(\lambda),\qquad C_i\sim \operatorname{Exp}(\lambda)$ even though $T_i$ and $C_i$ are independent. Then the censoring also depends on $\lambda$.

The full likelihood becomes

$$
L(\lambda)
=
\prod_{i=1}^n
\left[
f_T(y_i;\lambda)S_C(y_i;\lambda)
\right]^{\delta_i}
\left[
S_T(y_i;\lambda)f_C(y_i;\lambda)
\right]^{1-\delta_i}.
$$

Since $f_T(y;\lambda)=\lambda e^{-\lambda y},S_T(y;\lambda)=e^{-\lambda y},$

we get

$$
L(\lambda)
=
\lambda^n
\exp\left(-2\lambda\sum_{i=1}^n y_i\right).
$$

Thus the full-likelihood MLE is

$$
\widehat{\lambda}
=
\frac{n}{2\sum_{i=1}^n y_i}.
$$

But if we incorrectly ignore the censoring  , we would use only

$$
L_T(\lambda)
=
\prod_{i=1}^n
f_T(y_i;\lambda)^{\delta_i}
S_T(y_i;\lambda)^{1-\delta_i}
=
\lambda^{\sum_i\delta_i}
\exp\left(-\lambda\sum_i y_i\right),
$$

which gives

$$
\widehat{\lambda}
=
\frac{\sum_{i=1}^n\delta_i}
{\sum_{i=1}^n y_i}.
$$

These two estimators are generally different. 

---

## Problem 2
```{r,echo=FALSE}
print_no_call <- function(x) {
  out <- capture.output(print(x))
  call_lines <- grep("^Call:", out)
  if (length(call_lines) > 0) {
    drop <- integer()
    for (i in call_lines) {
      j <- i
      repeat {
        drop <- c(drop, j)
        j <- j + 1
        if (j > length(out) || out[j] == "") break
      }
    }
    out <- out[-unique(drop)]
  }
  cat(out, sep = "\n")
}
```


```{r problem2-load}
# Load the file Leuk.csv into R 
Leuk <- read.csv("Leuk.csv")
# 0 = Placebo, 1 = 6-MP
Leuk$drug_group <- factor(
  Leuk$drug,
  levels = c(0, 1),
  labels = c("Placebo", "6-MP")
)
# 0 = Female, 1 = Male
Leuk$sex_group <- factor(
  Leuk$sex,
  levels = c(0, 1),
  labels = c("Female", "Male")
)
```

```{r}
# Overview of the data
head(Leuk)
```

```{r}
summary(Leuk)
```

---

###  (a) *Do you observe any initial 'differences' between the two treatment groups?*

```{r problem2-a}
# Comparison of 2 groups
# time and logwbc are continuous
aggregate(
  cbind(time,logwbc) ~ drug_group,
  data = Leuk,
  FUN = function(x) c(mean = mean(x),sd = sd(x))
) 
```

```{r}
# cens and sex_group are categorical
# 0 = censoring, 1 = coming out of remission
with(Leuk, list(event = table(drug_group, cens),sex = table(drug_group, sex_group)))
```

- The 6-MP group has a **longer** average remission *time*, about 17.10 weeks, while the placebo group has an
average remission time of about 8.67 weeks. 
- **All** placebo patients came out of remission, but many 6-MP patients were **censored**, which suggests longer remission under 6-MP.
- The *sex* distribution is **balanced** between the two groups.

```{r problem2-surv-object}
leuk.survobj <- Surv(time = Leuk$time, event = Leuk$cens) # 创建生存对象
# 没有加号表示事件发生，加号表示右删失
leuk.survobj
```

---

### Kaplan-Meier Curves

```{r problem2-km-overall}
# obtain the overall K-M curve
leuk.kmoverall <- survfit(leuk.survobj ~ 1, data = Leuk)
# n.risk: 仍在观察中 survival: 生存概率 
print_no_call(summary(leuk.kmoverall))
```

```{r,echo=FALSE}
plot(leuk.kmoverall, conf.int = TRUE, col = 2)
title("Overall K-M curve")
```

```{r problem2-km-drug}
# obtain K-M curves for the treatment groups
leuk.kmdrug <- survfit(leuk.survobj ~ drug_group, data = Leuk)
knitr::kable(leuk.kmdrug$table)
```

```{r}
print_no_call(summary(leuk.kmdrug)) # summary(leuk.kmdrug)
```

```{r,echo=FALSE}
km23 <- summary(leuk.kmdrug, times = 23)
km23_table <- data.frame(
  group = sub("^drug_group=", "", km23$strata),
  time = km23$time,
  risk = km23$n.risk,
  event = km23$n.event,
  surv = km23$surv,
  se = km23$std.err,
  lower_CI = km23$lower,
  upper_CI = km23$upper
)
```

```{r}
knitr::kable(km23_table, digits = 3) # summary(leuk.kmdrug, times = 23)
```

```{r,echo=FALSE}
plot(leuk.kmdrug, col = 2:3, lty = 1:2,
     xlab = "Remission times",
     ylab = "Proportion in remission")
legend("topright", c("Placebo", "6-MP"), lty = 1:2, col = 2:3)
title("K-M curves for the two treatment groups")
```


###  (b) *What are these curves telling us?*

- The K-M curves show that patients receiving 6-MP generally remain in remission **longer** than patients receiving placebo. 
- At week 23, the estimated survivor probability is about 0 for placebo and about 0.45 for 6-MP.

###  (c) *Now plot the survival curves by sex and by logwbc.*

(*logwbc is continuous, should we group it ?*)

```{r problem2-c, echo=FALSE}
leuk.kmsex <- survfit(leuk.survobj ~ sex_group, data = Leuk)
plot(leuk.kmsex, col = 4:5, lty = 1:2,
     xlab = "Remission times",
     ylab = "Proportion in remission")
legend("topright", c("Female", "Male"), lty = 1:2, col = 4:5)
title("K-M curves by sex")
# logwbc is continuous, should we group it ?
leuk.kmlogwbc <- survfit(leuk.survobj ~ logwbc, data = Leuk)
plot(leuk.kmlogwbc, col = rainbow(length(leuk.kmlogwbc$strata)), lty = 1,
     xlab = "Remission times",
     ylab = "Proportion in remission")
title("K-M curves by observed logwbc values(no grouping)")
Leuk$logwbc_group <- cut(
  Leuk$logwbc,
  breaks = quantile(Leuk$logwbc, probs = c(0, 0.5, 1), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Low logwbc", "High logwbc")
)
leuk.logwbc_group <- survfit(leuk.survobj ~ logwbc_group, data = Leuk)
plot(leuk.logwbc_group, col = rainbow(length(leuk.logwbc_group$strata)), lty = 1,
     xlab = "Remission times",
     ylab = "Proportion in remission")
legend("topright", levels(Leuk$logwbc_group), col = rainbow(length(levels(Leuk$logwbc_group))), lty = 1)
title("K-M curves by observed logwbc values(grouped)")
```

---

### Log-Rank Tests

(*to assess whether patients in the 6-MP group remaining generally in remission longer*)

```{r problem2-logrank-drug}
logrank.drug <- survdiff(leuk.survobj ~ drug_group, data = Leuk)
print_no_call(logrank.drug)
```

###  (d) *What do you conclude? What are the test-statistic values for the log-rank statistic and its approximation?*

- The drug log-rank test gives a **chi-square statistic** of about 16.8 with 1 degree of freedom and **p-value** about $4e-05$.
- There is strong evidence that the remission distributions **differ** between placebo and 6-MP.

###  (e) *Perform now the log-rank test on sex and logwbc. What are your conclusions? Was it appropriate to use the log-rank test for these variables?*

```{r problem2-logrank-sex-logwbc}
# log-rank tests on sex
logrank.sex <- survdiff(leuk.survobj ~ sex_group, data = Leuk)
print_no_call(logrank.sex)
```

```{r}
# log-rank tests on continuous logwbc
#logrank.logwbc <- survdiff(leuk.survobj ~ logwbc, data = Leuk) 

# log-rank tests on grouped logwbc
logrank.logwbc <- survdiff(leuk.survobj ~ logwbc_group, data = Leuk) 
print_no_call(logrank.logwbc)
```

- The sex log-rank test is *not* significant, with **chi-square** about 0.6 and **p-value** about 0.5.
- The log-rank test is *not* really appropriate for raw continuous logwbc so that we **grouped** logwbc into 2 groups for log-rank tests.
- After grouping the continuous variable `logwbc` into low and high groups, the log-rank test shows a highly **significant** difference between the two survival curves.

- The low-logwbc group has 11 observed events, much fewer than the
21.39 events expected under the null hypothesis of equal survival curves. 

- The high-logwbc group has 19 observed events, much more than the
8.61 expected events.

- The log-rank **chi-square** statistic is 20.9 with 1 degree of freedom, and the
**p-value** is about \(5 \times 10^{-6}\). Therefore, we **reject** the null hypothesis
and conclude that remission times **differ significantly** between the two
logwbc groups.

- Patients with higher logwbc tend to come out of remission
earlier, while patients with lower logwbc tend to remain in remission longer.


---

### Cox Proportional Hazards Models

*To assess the effect of 6-MP on the hazard of coming out of remission, we begin by only including drug into the Cox PH model (i.e. fitting an unadjusted Cox PH model).*

We fit the Cox proportional hazards model

$$
h(t \mid \text{drug}) = h_0(t)\exp(\beta \cdot \text{drug}),
$$

where:

- $h(t \mid \text{drug})$ is the hazard of relapse at time $t$
- $h_0(t)$ is the baseline hazard for the placebo group
- $\text{drug}=0$ represents the placebo group, $\text{drug}=1$ represents the 6-MP treatment group
- $\beta$ measures the treatment effect

```{r problem2-cox-drug}
leuk.ph <- coxph(leuk.survobj ~ drug, data = Leuk)
print_no_call(summary(leuk.ph))
```

###  (f) *Do you understand the output? How do you interpret the drug effect? What hypothesis is the likelihood ratio and Wald tests in the output testing?*

The fitted model is

$$
\hat h(t \mid \text{drug}) = h_0(t)\exp(-1.5721 \cdot \text{drug}).
$$

Thus, the estimated hazard ratio for 6-MP versus placebo is

$$
\exp(\hat\beta) = \exp(-1.5721) = 0.2076.
$$

- The hazard ratio is about **0.208**, meaning that the relapse hazard for patients receiving 6-MP is approximately 20.8% of that for patients receiving placebo.

- The p-value for the drug coefficient is $p = 0.000138,$ which is much smaller than 0.05. Therefore, the treatment effect is statistically **significant**.

- The 95% confidence interval for the hazard ratio is $(0.0925,\ 0.4659),$ which is entirely **below** 1. This also indicates that 6-MP significantly **lowers** the risk of relapse compared with placebo.

- Overall, the Cox model suggests that patients treated with 6-MP tend to remain in remission **longer** than patients in the placebo group.

- The likelihood ratio and Wald tests are testing the null hypothesis that the *drug coefficient is zero*.


###  (g) *Based on the variables provided in Leuk, come up with a "best" fitting Cox PH model. Is there a drug by logwbc interaction? Interpret the results from your best/final model.*

```{r problem2-cox-models}
leuk.ph.drug.logwbc <- coxph(leuk.survobj ~ drug + logwbc, data = Leuk) 
leuk.ph.full <- coxph(leuk.survobj ~ drug + sex + logwbc, data = Leuk) 
leuk.ph.int <- coxph(leuk.survobj ~ drug * logwbc + sex, data = Leuk)
anova(leuk.ph.full, leuk.ph.int, test = "Chisq") #  检验交互项
```

```{r}
print_no_call(summary(leuk.ph.drug.logwbc)) # drug + logwbc
```

```{r}
print_no_call(summary(leuk.ph.full)) # drug + sex + logwbc
```

```{r}
print_no_call(summary(leuk.ph.int)) # drug * logwbc + sex
```

```{r}
AIC(leuk.ph,leuk.ph.drug.logwbc,leuk.ph.full,leuk.ph.int) # 比较各个模型的AIC
```

- The drug-by-logwbc interaction is *not* statistically significant, with a p-value of about 0.464.
- We also compared the additive Cox model with the interaction model using a likelihood ratio test. The test statistic is 0.5286 with 1 degree of freedom, and the **p-value** is 0.4672, which is much larger than 0.05. 
- Among the fitted Cox models, the model **including drug and continuous logwbc** has the smallest AIC, so it is selected as the final model.
- In this final model, the `hazard ratio` for 6-MP versus placebo is 0.250, meaning that 6-MP reduces the relapse hazard by about 75% after adjusting for logwbc.

- For logwbc, the `hazard ratio` is 5.424 per one-unit increase, suggesting that higher logwbc is strongly associated with earlier relapse.

---

###  (h) *Can you figure out how to plot the predicted survival curve for a person who is on 6-MP and has a logwbc value of 4.00? You may want to look at the help file for the R command survfit.*

```{r problem2-predicted-survival}
new.patient <- data.frame(drug = 1, logwbc = 4.00)
new.patient.surv <- survfit(leuk.ph.drug.logwbc, newdata = new.patient)
print_no_call(new.patient.surv)
```

```{r,echo=FALSE}
plot(new.patient.surv, conf.int = TRUE, col = 2,
     xlab = "Remission time",
     ylab = "Predicted proportion in remission")
```

---

### Assessing the Proportional Hazards Assumption

*To assess the validity of the PH assumption for the drug variable, begin by plotting the complementary log-log curves for the Placebo and 6-MP groups.*

```{r,echo=FALSE}
# To assess the validity of the PH assumption for the drug variable
plot(survfit(leuk.survobj ~ drug_group, data = Leuk),
     col = 2:3, lty = 1:2, fun = "cloglog",
     xlim = c(1, 40), ylim = c(-3, 3),
     xlab = "Time of remission",
     ylab = "log(-log(S(t)))")
legend("topleft", c("Placebo", "6-MP"), lty = 1:2, col = 2:3)
```

```{r}
cox.zph(leuk.ph) # using the default K-M transformation of time
```

###  (i) *What do you conclude?*

- The complementary log-log curves do *not* show a major violation for drug. 
- The Schoenfeld residual test gives a p-value about 0.88 for drug, so there is *no* evidence against the proportional hazards assumption for the drug variable in the unadjusted Cox model.

###  (j) *Test whether the PH assumption is satisfied for sex and logwbc.*

```{r,echo=FALSE}
# Test whether the PH assumption is satisfied for sex
plot(survfit(leuk.survobj ~ sex_group, data = Leuk),
     col = 4:5, lty = 1:2, fun = "cloglog",
     xlim = c(1, 40), ylim = c(-3, 3),
     xlab = "Time of remission",
     ylab = "log(-log(S(t)))")
legend("topleft", c("Female", "Male"), lty = 1:2, col = 4:5)
```

```{r}
cox.zph(leuk.ph.full)
```

(*For `logwbc`, which is a continuous variable, the complementary log-log plot should be based on grouped values rather than treating every observed value as a separate group.*)


```{r,echo=FALSE}
# Test whether the PH assumption is satisfied for logwbc
Leuk$logwbc_group <- cut(
  Leuk$logwbc,
  breaks = quantile(Leuk$logwbc, probs = c(0, 0.5, 1), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Low logwbc","High logwbc")
)
leuk.cloglog.logwbc <- survfit(leuk.survobj ~ logwbc_group, data = Leuk)
plot(leuk.cloglog.logwbc,
     col = rainbow(length(leuk.cloglog.logwbc$strata)), lty = 1, fun = "cloglog",
     xlim = c(1, 40), ylim = c(-3, 3),
     xlab = "Time of remission",
     ylab = "log(-log(S(t)))")
legend("topleft", levels(Leuk$logwbc_group),
       col = rainbow(length(leuk.cloglog.logwbc$strata)), lty = 1)
```

```{r}
# check the selected Cox model with drug and logwbc
cox.zph(leuk.ph.drug.logwbc) 
```

- In the full Cox model including `drug`, `sex`, and `logwbc`, the Schoenfeld residual test gives a **p-value** for sex about 0.02, suggesting a possible **violation** of the proportional hazards assumption for sex.
- In the final selected Cox model with `drug` and `logwbc`, the Schoenfeld residual test does *not* show evidence of PH violation for either variable. The **p-values** are about 0.99 for drug, 0.84 for logwbc, and 0.98 for the global test.
- As for the full model where `sex` is included, a reasonable approach is to stratify by sex. 
- In the stratified Cox model, the PH tests for `drug` and `logwbc` are *not* significant, so the stratified model also appears appropriate.

(*If the PH assumption is not satisfied, re-fit the Cox model but now stratifying by the variable not satisfying the PH assumption.*)

```{r}
leuk.sc <- coxph(leuk.survobj ~ drug + logwbc + strata(sex), data = Leuk)
print_no_call(summary(leuk.sc))
cox.zph(leuk.sc)
```

---

## Problem 3

```{r problem3-setup}
# load the data
lungdat <- survival::lung
# 0 = censored, 1 = dead
lungdat$status <- lungdat$status - 1 # creating 0/1 event indicator
# ECOG 0/ECOG 1/ECOG 2+
lungdat$ph.ecogcat <- ifelse(lungdat$ph.ecog >= 2, 2, lungdat$ph.ecog)
# wt.loss = weight loss in last six months; positive values mean weight loss
lungdat$wtlossyn <- ifelse(lungdat$wt.loss > 0, 1, 0)
head(lungdat)
```

###  (a) *What does the overall survival distribution look like?*

```{r problem3-km-overall}
lung.survobj <- Surv(time = lungdat$time, event = lungdat$status) # 创建生存对象
lung.kmoverall <- survfit(lung.survobj ~ 1, data = lungdat) # 包括所有病人，不分组
print_no_call(lung.kmoverall)
```

```{r}
print_no_call(summary(lung.kmoverall, times = c(180, 365, 730))) # 挑3个时间点
```

```{r problem3-km-overall-plot, echo=FALSE}
plot(lung.kmoverall, conf.int = TRUE, col = 2,
     xlab = "Survival time (days)",
     ylab = "Survival probability")
title("Overall K-M curve for the lung cancer data")
```

- The overall survival curve **decreases** steadily over time. 
- The **median** survival time is about 310 days
- The estimated **survival probability** at one year is roughly 0.41

---

### (b) *Which of these four distributions is preferred? Why?*

```{r problem3-parametric-models}
ex <- flexsurvreg(Surv(time, status) ~ 1, data = lungdat, dist = "exponential")
we <- flexsurvreg(Surv(time, status) ~ 1, data = lungdat, dist = "weibull")
go <- flexsurvreg(Surv(time, status) ~ 1, data = lungdat, dist = "gompertz")
gg <- flexsurvreg(Surv(time, status) ~ 1, data = lungdat, dist = "gengamma")
```

```{r,echo=FALSE}
parametric_models <- list(
  Exponential = ex,
  Weibull = we,
  Gompertz = go,
  GenGamma = gg
)
model_compare <- data.frame(
  distribution = names(parametric_models),
  logLik = sapply(parametric_models, function(x) x$loglik),
  AIC = sapply(parametric_models, function(x) x$AIC),
  row.names = NULL
)
model_compare <- model_compare[order(model_compare$AIC), ]
knitr::kable(model_compare, digits = 3)
```

- The `GenGamma` distribution has the highest log-likelihood, but the log-likelihood of the `Weibull` distribution is very close to it.
- A **smaller** AIC indicates a better balance between model fit and model complexity. As a result, the `Weibull` distribution provides the best fit among the four parametric models.

### (c) *What do you conclude from this figure?*

```{r problem3-parametric-survival-plot, echo=FALSE}
plot(lung.kmoverall, conf.int = FALSE, mark.time = TRUE,
     xlab = "Time (days)", ylab = "Survival",col = "black", lwd = 2)
lines(ex, col = "purple", ci = FALSE, lwd = 2)
lines(we, col = "blue", ci = FALSE, lwd = 2)
lines(go, col = "green", ci = FALSE, lwd = 2)
lines(gg, col = "red", ci = FALSE, lwd = 2)
legend("topright",
       lty = c(1, 1, 1, 1, 1),
       lwd = c(2, 2, 2, 2, 2),
       col = c("black", "purple", "blue", "green", "red"),
       c("KM", "Exp", "Weib", "Gomp", "GenGam"))
```

- The `exponential` model appears to 
provide a poorer fit to the Kaplan-Meier curve. 
- It tends to underestimate survival at earlier times and overestimate survival at later times, 
suggesting that the constant hazard assumption may be too restrictive.

- The `Weibull`, `Gompertz`, and `generalized gamma` models fit the Kaplan-Meier curve much better. 

---

### (d)*To determine the hazard function for the fitted exponential model, you can run the command `summary(ex, type="hazard",ci=TRUE,tidy=FALSE)`. Can you figure out how to plot the estimated hazard functions for these four distributions on the same figure?*

```{r problem3-exponential-hazard}
# hazard summary for the exponential model
# the full output is quite long so we only show 3 time points
summary(ex, type = "hazard",t=c(50,100,150),ci = TRUE, tidy = FALSE)
```

```{r problem3-hazard-plot, echo=FALSE}
time_grid <- seq(1, max(lungdat$time, na.rm = TRUE), length.out = 500)
hazard_data <- do.call(
  rbind,
  lapply(names(parametric_models), function(nm) {
    hazard_summary <- summary(
      parametric_models[[nm]],
      type = "hazard",
      t = time_grid,
      ci = FALSE,
      tidy = TRUE
    )
    data.frame(
      distribution = nm,
      time = hazard_summary$time,
      hazard = hazard_summary$est
    )
  })
)
ggplot(hazard_data, aes(x = time, y = hazard, color = distribution)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (days)", y = "Estimated hazard",
       color = "Distribution") +
  theme_minimal()
```

- The `exponential` model has a **flat** hazard by construction. 
- The other models allow the hazard to change over time, which is more realistic if the risk of death is not constant throughout follow-up.

###  (e) *Interpret the output.*

```{r problem3-weibull-ecog}
lungdat_ecog <- subset(lungdat, !is.na(ph.ecogcat))
we.reg <- flexsurvreg(
  Surv(time, status) ~ factor(ph.ecogcat),
  data = lungdat_ecog,
  dist = "weibull"
)
print_no_call(we.reg)
```

```{r,echo=FALSE}
we_shape <- we.reg$res["shape", "est"]
ecog_coef_names <- grep("ph\\.ecogcat", rownames(we.reg$res), value = TRUE)
ecog_effects <- data.frame(
  term = ecog_coef_names,
  beta = we.reg$res[ecog_coef_names, "est"],
  acceleration_factor = exp(we.reg$res[ecog_coef_names, "est"]),
  weibull_PH_hazard_ratio = exp(-we_shape * we.reg$res[ecog_coef_names, "est"]),
  row.names = NULL
)
knitr::kable(ecog_effects, digits = 3)
```

- The Weibull AFT model uses `ECOG 0` as the reference group.

- The estimated acceleration factor for `ECOG 1` is 0.767, suggesting **shorter** survival compared with ECOG 0. However, its 95% confidence interval includes 1, so this effect is not clearly significant at the 5% level.

- The estimated acceleration factor for `ECOG 2` is 0.505, with a 95% confidence interval entirely below 1, indicating significantly **shorter** survival compared with ECOG 0.

- The Weibull shape parameter is 1.356, which is greater than 1. This suggests that the baseline hazard **increases** over time.

###  (f) *How would you interpret this output with regard to accelerating life and proportional hazards?*

- Since the Weibull model can also be interpreted in proportional hazards form, the corresponding hazard ratio is given by $\exp(-\hat\alpha\hat\beta)$, where $\hat\alpha$ is the Weibull shape parameter and $\hat\beta$ is the AFT coefficient.

- `ECOG 1` has an estimated mortality hazard about 1.43 times that of ECOG 0, while `ECOG 2` has an estimated mortality hazard about 2.52 times that of ECOG 0.

- **Worse** ECOG performance status is associated with shorter survival time and higher mortality risk.

###  (g) *What do you conclude from looking at this plot?*

```{r problem3-cloglog-ecog, echo=FALSE}
lung.kmph.ecog <- survfit(lung.survobj ~ factor(ph.ecogcat), data = lungdat)
plot(lung.kmph.ecog, col = 2:4, lty = 1:3, fun = "cloglog", log = "x",
     xlab = "t on log-scale",
     ylab = "log(-log(S(t)))")
legend("topleft", c("ECOG 0", "ECOG 1", "ECOG 2+"), col = 2:4, lty = 1:3)
```

- The complementary log-log curves show clear separation between the ECOG groups. 
- The `ECOG 2+` group tends to have a **higher** curve, suggesting a higher cumulative hazard and shorter survival compared with the `ECOG 0` and `ECOG 1` groups.
- The curves are *not* perfectly straight or perfectly parallel on the log-time scale, which suggests that the common-shape Weibull PH/AFT structure may only be an approximate working model. 
- The non-parallel pattern indicates that ECOG may also affect the shape of the Weibull hazard function, not only the scale of survival time.
- Since the plot is based on KM step functions and the sample size becomes smaller at later times, some departures from parallelism are expected. 
- The plot does not completely rule out the Weibull model, but it motivates checking a more flexible Weibull model in which the shape parameter is allowed to depend on ECOG.

###  (h) *What impact does ECOG have on the shape of the Weibull regression model just fitted?*

```{r problem3-ecog-shape}
# ECOG 影响 scale 同时还影响 shape
we.reg.shape.ecog <- flexsurvreg(
  Surv(time, status) ~ factor(ph.ecogcat),
  anc = list(shape = ~ factor(ph.ecogcat)),
  data = lungdat_ecog,
  dist = "weibull"
)
ecog_shape_compare <- data.frame(
  model = c("Common shape", "Shape depends on ECOG"),
  logLik = c(we.reg$loglik, we.reg.shape.ecog$loglik),
  AIC = c(we.reg$AIC, we.reg.shape.ecog$AIC)
)
knitr::kable(ecog_shape_compare, digits = 3)
```

```{r}
print_no_call(we.reg.shape.ecog)
```

- The model allowing the Weibull `shape` parameter to depend on ECOG has a **slightly higher** log-likelihood than the common-shape model, but its AIC is **larger**. This means that the improvement in model fit is *not* large enough to justify the additional shape parameters.

- The estimated shape effects for ECOG 1 and ECOG 2+ are also not clearly significant, since their confidence intervals include 1 after exponentiation. 
- There is no strong evidence that ECOG has an important effect on the `shape` of the Weibull hazard function.
- The simpler common-shape Weibull model is preferred. ECOG mainly appears to affect the `scale` of survival time, with worse ECOG performance associated with shorter survival, rather than changing the `shape` of the Weibull survival distribution.

###  (i) *What impact does weight loss in the last six months (`wtlossyn` variable) have on the risk and the shape of the Weibull regression model just fitted?*

```{r problem3-weight-loss-models}
lungdat_wt <- subset(lungdat, !is.na(ph.ecogcat) & !is.na(wtlossyn))
# ECOG 和 wtlossyn 只影响 scale
we.reg.wt <- flexsurvreg(
  Surv(time, status) ~ factor(ph.ecogcat) + factor(wtlossyn),
  data = lungdat_wt,
  dist = "weibull"
)
# ECOG 和 wtlossyn 影响 scale，同时 wtlossyn 还影响 shape
we.reg.wt.shape <- flexsurvreg(
  Surv(time, status) ~ factor(ph.ecogcat) + factor(wtlossyn),
  anc = list(shape = ~ factor(wtlossyn)),
  data = lungdat_wt,
  dist = "weibull"
)
wt_compare <- data.frame(
  model = c("Weight loss affects scale", "Weight loss affects scale and shape"),
  logLik = c(we.reg.wt$loglik, we.reg.wt.shape$loglik),
  AIC = c(we.reg.wt$AIC, we.reg.wt.shape$AIC)
)
knitr::kable(wt_compare, digits = 3)
```

```{r}
print_no_call(we.reg.wt) # weight loss affects scale
print_no_call(we.reg.wt.shape) # weight loss affects scale and shape
```

- Here `wtlossyn = 1` means the patient lost weight in the last six months, while `wtlossyn = 0` means the patient did not lose weight or gained weight.
- After adjusting for ECOG performance score, `wtlossyn` does *not* appear to have a strong effect on the `scale` of the Weibull AFT model. 
- In the common-shape model, the acceleration factor for `wtlossyn` is about 1.115, and its 95% confidence interval includes 1. This suggests that weight loss does *not* substantially shift the overall survival time `scale`.
- When `wtlossyn` is allowed to affect the Weibull `shape` parameter, the AIC **increases slightly** from 2119.138 to 2120.346. The estimated `shape` ratio for `wtlossyn` is about 0.880, with a 95% confidence interval from 0.668 to 1.161. This does not provide clear evidence that weight loss affects the `shape` of the hazard function.
- There is no strong evidence that weight loss in the last six months has an important effect on either the survival time `scale` or the Weibull hazard `shape`.

---

###  (j) *Can you figure out how to plot the predicted survival curve for an asymptomatic, advanced lung cancer patient who did not lose weight in the last six months?*

```{r problem3-predicted-survival}
new.lung.patient <- data.frame(ph.ecogcat = 0,wtlossyn = 0) # 预测对象
pred.time <- seq(1, max(lungdat_wt$time, na.rm = TRUE), length.out = 200)
pred.surv <- summary(
  we.reg.wt, # common-shape model with ECOG and weight loss affecting scale
  type = "survival",
  t = pred.time,
  newdata = new.lung.patient,
  ci = TRUE,
  tidy = TRUE
)
# est 表示生存概率
head(pred.surv)
```

```{r problem3-predicted-survival-plot, echo=FALSE}
ggplot(pred.surv, aes(x = time, y = est)) +
  geom_ribbon(aes(ymin = lcl, ymax = ucl), alpha = 0.2, fill = "grey70") +
  geom_line(color = 2, linewidth = 0.8) +
  labs(x = "Time (days)", y = "Predicted survival probability") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_minimal()
```

- The predicted survival curve is obtained for an asymptomatic advanced lung cancer patient with `ECOG 0` and `wtlossyn = 0`. 
- The model used for prediction is the common-shape Weibull model with ECOG and `wtlossyn` affecting the `scale` parameter.
- The fitted survival probability starts close to 1 at early times and decreases gradually over time. 
- The predicted median survival time is roughly around 410 days, where the survival curve crosses 0.5. 
- By around 1000 days, the predicted survival probability is about 0.09. 
- The shaded region represents the 95% confidence interval, which becomes wider at later times, indicating greater uncertainty in long-term survival prediction.

