---
title: "Mathematical practice final exam 2024"
date: '2024-07-08'
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r message = F, echo = F}
library(tidyverse)
# install.packages("Ecdat")
#library(Ecdat)
```

## 1. 
Solve the following system of equations using the `solve()` function:
$$
\left(\begin{array}{cccc}
9 & 4 & 12 & 2 \\
5 & 0 & 7 & 9 \\
2 & 6 & 8 & 0 \\
9 & 2 & 9 & 11
\end{array}\right) \times\left(\begin{array}{l}
x \\
y \\
z \\
t
\end{array}\right)=\left(\begin{array}{c}
7 \\
18 \\
1 \\
0
\end{array}\right)
$$
```{r}
A <- matrix(c(9, 4, 12, 2,
              5, 0, 7, 9,
              2, 6, 8, 0,
              9, 2, 9, 11),
            nrow = 4, byrow = TRUE)
b <- c(7, 18, 1, 0)
solution <- solve(A, b)
print(solution)
```


## 2. 
Execute the following lines which create two vectors of random integers which are chosen with replacement from the integers $0, 1, \dots , 999$. Both vectors have length 250. 

```{r}
xVec <- sample(0:999, 250, replace=T)
yVec <- sample(0:999, 250, replace=T)
```

(a) Create the vector $(y_2 - x_1, \cdots, y_n - x_{n-1}).$
```{r}
vec <- yVec[2:length(yVec)]-xVec[1:(length(xVec))-1]
print(vec)
```


(b) Pick out the values in yVec which are $> 600$.

```{r}
vec_b <- yVec[yVec > 600]
print(vec_b)
```


(c) What are the index positions in yVec of the values which are $> 600$? 

```{r}
which(yVec>600)
```


(d) Sort the numbers in the vector xVec in the order of increasing values in yVec. 

```{r}
vec_d <- xVec[order(yVec)]
print(vec_d)
```


(e) Pick out the elements in yVec at index positions $1, 4, 7, 10, 13, \cdots$

```{r}
vec_e <- yVec[seq(1, length(yVec), by = 3)]
print(vec_e)
```

## 3.
For this problem we'll use the (built-in) dataset state.x77.
```{r}
data(state)
state.x77 <- as_tibble(state.x77, rownames  = 'State')
```

a. Select all the states having an income less than 4300, and calculate the average income of these states.
```{r}
a <- state.x77 |> filter(Income < 4300)
mean(pull(a, Income))
```

b.	Sort the data by income and select the state with the highest income.
```{r}
state.x77 |> arrange(Income) |> tail(1)
state.x77 |> summarise(max(Income))
```

c.	Add a variable to the data frame which categorizes the size of population: $<= 4500$ is `S`, $ > 4500 $ is `L`.
```{r, include=FALSE}
state.x77<-state.x77 %>% 
  mutate(Size = ifelse(Population <= 4500, "S", "L"))
```

d.	Find out the average income and illiteracy of the two groups of states, distinguishing by whether the states are small or large.
```{r, include=FALSE}
d_summary <- state.x77 %>% 
  group_by(Size) %>% 
  summarise(
    avg_income = mean(Income),
    avg_Illiteracy = mean(Illiteracy)
  )
d_summary
```

## 4. 
a. Write a function to simulate `n` observations of $(X_1, X_2)$ which follow the uniform distribution over the square $[0, 1] \times [0, 1]$.
```{r}
simulate_uniform_points <- function(n) {
  data.frame(
    X1 = runif(n, 0, 1),
    X2 = runif(n, 0, 1)
  )
}
simulate_uniform_points(100)
```

b. Write a function to calculate the proportion of the observations that the distance between $(X_1, X_2)$ and the nearest edge is less than 0.25, and the proportion of them with the distance to the nearest vertex less than 0.25.
```{r}
calculate_proportions <- function(data) {
  # Distance to nearest edge
  d_edge <- pmin(data$X1, 1 - data$X1, data$X2, 1 - data$X2)
  prop_edge <- mean(d_edge < 0.25)

  # Distance to nearest vertex
  d1 <- sqrt(data$X1^2 + data$X2^2)                    # to (0, 0)
  d2 <- sqrt(data$X1^2 + (1 - data$X2)^2)              # to (0, 1)
  d3 <- sqrt((1 - data$X1)^2 + data$X2^2)              # to (1, 0)
  d4 <- sqrt((1 - data$X1)^2 + (1 - data$X2)^2)        # to (1, 1)
  d_vertex <- pmin(d1, d2, d3, d4)
  prop_vertex <- mean(d_vertex < 0.25)

  # Return result as named vector
  c(proportion_edge = prop_edge,
    proportion_vertex = prop_vertex)
}
calculate_proportions(simulate_uniform_points(100))
```

## 5.

To estimate $\pi$ with a Monte Carlo simulation, we draw the unit circle inside the unit square, the ratio of the area of the circle to the area of the square will be $\pi / 4$. Then shot $K$ arrows at the square, roughly $K * \pi / 4$ should have fallen inside the circle. So if now you shoot $N$ arrows at the square, and $M$ fall inside the circle, you have the following relationship $M = N * \pi / 4$. You can thus compute $\pi$ like so: $\pi = 4 * M / N$. The more arrows $N$ you throw at the square, the better approximation of $\pi$ you'll have.

```{r}
n <- 10000

set.seed(1)
points <- tibble("x" = runif(n), "y" = runif(n))
```

Now, to know if a point is inside the unit circle, we need to check whether $x^2 + y^2 < 1$. Let's add a new column to the points tibble, called `inside` equal to `1` if the point is inside the unit circle and `0` if not:

```{r}
points <- points |> 
    mutate(inside = map2_dbl(.x = x, .y = y, ~ifelse(.x**2 + .y**2 < 1, 1, 0))) |> 
    rowid_to_column("N")
```

a. Compute the estimation of $\pi$ at each row, by computing the cumulative sum of the 1's in the `inside` column and dividing that by the current value of `N` column:

```{r}
library(tidyverse)


points_with_pi <- points %>%
  mutate(
    cum_inside = cumsum(inside),   
    pi_estimate = 4 * cum_inside / N  
  )
tail(points_with_pi)
```

b. Plot the estimates of $\pi$ against `N`.

```{r}
ggplot(points_with_pi, aes(x = N, y = pi_estimate)) +
  geom_line() +                      
  geom_hline(yintercept = pi, color = "red", linetype = "dashed") +  
  labs(title = "Estimation of π using Monte Carlo Simulation",
       x = "Number of Points (N)",
       y = "Estimated π") +
  theme_minimal()
```

## 6. 
Mortality rates per 100,000 from male suicides for a number of age groups and a number of countries are given in the following data frame. 
```{r}
suicrates <- tibble(Country = c('Canada', 'Israel', 'Japan', 'Austria', 'France', 'Germany',
'Hungary', 'Italy', 'Netherlands', 'Poland', 'Spain', 'Sweden', 'Switzerland', 'UK', 'USA'), 
Age25.34 = c(22,  9, 22, 29, 16, 28, 48,  7,  8, 26,  4, 28, 22, 10, 20), 
Age35.44 = c(27, 19, 19, 40, 25, 35, 65,  8, 11, 29,  7, 41, 34, 13, 22), 
Age45.54 = c(31, 10, 21, 52, 36, 41, 84, 11, 18, 36, 10, 46, 41, 15, 28), 
Age55.64 = c(34, 14, 31, 53, 47, 49, 81, 18, 20, 32, 16, 51, 50, 17, 33), 
Age65.74 = c(24, 27, 49, 69, 56, 52, 107, 27, 28, 28, 22, 35, 51, 22, 37))
```

a. Transform `suicrates` into *long* form.
```{r}
library(tidyverse)

# 转换为 long 格式
suicrates_long <- suicrates %>%
  pivot_longer(
    cols = starts_with("Age"),         
    names_to = "AgeGroup",             
    values_to = "Rate"                 
  )

head(suicrates_long)
```

b. Construct side-by-side box plots for the data from different age groups, and comment on what the graphic tells us about the data. 
```{r}
ggplot(suicrates_long, aes(x = AgeGroup, y = Rate)) +
  geom_boxplot() +
  labs(title = "Male Suicide Rates by Age Group",
       x = "Age Group",
       y = "Suicide Rate per 100,000") +
  theme_minimal() +
  coord_flip() 
```

