---
title: "Homework 4: Diffusion of Tetracycline"
author: "Zhang Zile,3230104237"
output: pdf_document
---
```{r,echo=FALSE}
options(warn = -1)  
```
We continue examining the diffusion of tetracycline among doctors in Illinois in the early 1950s, building on our work in lab 6.  You will need the data sets `ckm_nodes.csv` and `ckm_network.dat` from the labs.

1. Clean the data to eliminate doctors for whom we have no adoption-date information, as in the labs.  Only use this cleaned data in the rest of the assignment.
```{r, message=F}
library(tidyverse)
ckm_nodes <- read.csv('data/ckm_nodes.csv',header=T, sep = ",") 
noinfor <- which(is.na(ckm_nodes$adoption_date))
ckm_nodes <- ckm_nodes[-noinfor, ]
ckm_network <- read.table('data/ckm_network.dat')
ckm_network <- ckm_network[-noinfor, -noinfor]
```

2. Create a new data frame which records, for every doctor, for every month, whether that doctor began prescribing tetracycline that month, whether they had adopted tetracycline before that month, the number of their contacts who began prescribing strictly _before_ that month, and the number of their contacts who began prescribing in that month or earlier.  Explain why the dataframe should have 6 columns, and 2125 rows. 
```{r}
doctors <- rownames(ckm_nodes)
df <- data.frame(
  doctor = rep(doctors, each = 17),
  month = rep(1:17, times = length(doctors))
)
df <- df |> mutate(that_month = as.numeric(ckm_nodes[doctor,2] == month),
                   that_before = as.numeric(ckm_nodes[doctor,2] < month))
f <- function(x) {
    return(sum(ckm_nodes[ckm_network[as.numeric(x[1]), ] == 1, 2] < as.numeric(x[2])))
}
df <- df |>  mutate(contacts_str_before = apply(df, 1, f))
f <- function(x){
    return(sum(ckm_nodes[ckm_network[as.numeric(x[1]),] == 1, 2] <= as.numeric(x[2])))
}

df <- df  |>  mutate(contacts_in_before = apply(df, 1, f))
```
* There are 6 variables, so the dataframe has 6 columns.

* For every doctor, for every month $\Rightarrow$ 125 doctors $\times$ 17 months  = 2125, so the datafram has 2125 rows.
3.  Let

```{=latex}
    \begin{equation}
    \begin{split}
    p_k & = \Pr(\text{A doctor starts prescribing tetracycline this month} \mid \\
     & \text{Number of doctor's contacts prescribing before this month}=k) 
     \end{split}
     \end{equation}
```

and

```{=latex}
    \begin{equation}
    \begin{split}
     q_k & = \Pr(\text{A doctor starts prescribing tetracycline this month} \mid \\ 
     & \text{Number of doctor's contacts prescribing this month}=k)
     \end{split}
     \end{equation}
```

   We suppose that $p_k$ and $q_k$ are the same for all months.
  
  a. Explain why there should be no more than $21$ values of $k$ for which we can estimate $p_k$ and $q_k$ directly from the data.
```{r}
max(apply(ckm_network, 1, sum))
```
* Because the maximum of the contacts of a doctor is 20, so the possible values of $k$ are from 0 to 20, the number of which is no more than $21$.
  b. Create a vector of estimated $p_k$ probabilities, using the data frame from (2).  Plot the probabilities against the number of prior-adoptee contacts $k$.
```{r}
df_clean <- df[!is.na(df$contacts_str_before), ]
 p.vec <- vector(mode = "numeric",length = 21)
    k.vec <- p.vec
    for(k in 0:20){
        idx <- df_clean$contacts_str_before == k
        k.vec[k+1] <- sum(idx)
         if(is.na(k.vec[k+1]) | k.vec[k+1] == 0) {
        p.vec[k+1] <- NA
    } else {
        dfk <- df_clean[idx, ]
        p1 <- sum(dfk$that_month == 1)
        p.vec[k+1] <- p1 / k.vec[k+1]
    }
    }
    k <- c(0:20)
    plot(k.vec,p.vec, xlab="num k", ylab="p")
```
  c. Create a vector of estimated $q_k$ probabilities, using the data frame from (2). Plot the probabilities against the number of prior-or-contemporary-adoptee contacts $k$.
```{r}
df_clean <- df_clean[!is.na(df_clean$contacts_in_before), ]
p.vec2 <- vector(mode = "numeric",length = 21)
    k.vec2 <- p.vec2
    for(k in 0:20){
        idx <- (df_clean$contacts_in_before - df_clean$contacts_str_before) == k
        k.vec2[k+1] <- sum(idx)
        if(is.na(k.vec2[k+1]) | k.vec2[k+1] == 0){
            p.vec2[k+1] <- NA
        } else{
        dfk <- df_clean[idx,]
        p1 <- sum(dfk$that_month == 1)
        p.vec2[k+1] <- p1/k.vec2[k+1]}
    }
    k <- c(0:20)
    plot(k.vec2 + k.vec, p.vec2, xlab="num k", ylab="q")
```

4. Because it only conditions on information from the previous month, $p_k$ is a little easier to interpret than $q_k$.  It is the probability per month that a doctor adopts tetracycline, if they have exactly $k$ contacts who had already adopted tetracycline.
  a. Suppose $p_k = a + bk$.  This would mean that each friend who adopts the new drug increases the probability of adoption by an equal amount.  Estimate this model by least squares, using the values you constructed in (3b).  Report the parameter estimates.
```{r}
    df.p <- data.frame(k = 0:20, p = p.vec)
    m.1 <- lm(p ~ k, data = df.p)
    summary(m.1)
```
```
  Estimate    Std.Error    t   Pr(>|t|)    
a  0.056430   0.010564   5.342 0.000104 
b -0.003908   0.001089  -3.590 0.002956 
```
  b. Suppose $p_k = e^{a+bk}/(1+e^{a+bk})$.  Explain, in words, what this model would imply about the impact of adding one more adoptee friend on a given doctor's probability of adoption.  (You can suppose that $b > 0$, if that makes it easier.)  Estimate the model by least squares, using the values you constructed in (3b).
* It is a logistic curve. Suppose $b>0$. As $k$ grows, the initial stage of growth is approximately exponential (geometric); then, as saturation begins, the growth slows to linear (arithmetic), and at maturity, growth stops.
```{r}
f <- function(k,a,b){
        return(exp(a+b*k)/(1+exp(a+b*k)))
    }
    m.2 <- nls(p ~ f(k, a, b), data = df.p, start = list(a = 0, b = -0.2))
    summary(m.2)
```
  c. Plot the values from (3b) along with the estimated curves from (4a) and (4b).  (You should have one plot, with $k$ on the horizontal axis, and probabilities on the vertical axis .)  Which model do you prefer, and why?

```{r}
m.3 <- lm(p.log ~ k, df.p  |> 
                  mutate(p.log = ifelse(p==0, log(0.0001/(1-0.0001)), log(p/(1-p)))))
m1 = predict(m.1, newdata=data.frame(k=c(0:20)))
    m2 = predict(m.2, newdata=data.frame(k=c(0:20)))
    df.p <- mutate(df.p, linear = m1, logistic = m2)
    t = predict(m.3, newdata=data.frame(k=c(0:20)))
    m3 = exp(t)/(1+exp(t))
    df.p <- mutate(df.p, linear = m1, logistic = m2, logistic.2 = m3)
    df.tidy <- df.p  |>  gather(model, res, -k)  |>  na.omit()
    df.tidy  |>  ggplot() + geom_point(aes(x = k, y = res, color = model), size = 3) +
          geom_line(aes(x = k, y = res, color = model), linewidth = 1) +
          labs(y = "Probability",
               title = "Prediction of linear and logistic models")
```
*  The linear model is the worst. I use two methods to establish the logistic model and I think green line `logistic` using function `nls()` looks better on the whole. Because `logistic.2` seems a little overfitting in the tail part.

_For quibblers, pedants, and idle hands itching for work to do_: The $p_k$ values from problem 3 aren't all equally precise, because they come from different numbers of observations.  Also, if each doctor with $k$ adoptee contacts is independently deciding whether or not to adopt with probability $p_k$, then the variance in the number of adoptees will depend on $p_k$.  Say that the actual proportion who decide to adopt is $\hat{p}_k$.  A little probability (exercise!) shows that in this situation, $\mathbb{E}[\hat{p}_k] = p_k$, but that $\mathrm{Var}[\hat{p}_k] = p_k(1-p_k)/n_k$, where $n_k$ is the number of doctors in that situation.  (We estimate probabilities more precisely when they're really extreme [close to 0 or 1], and/or we have lots of observations.)  We can estimate that variance as $\hat{V}_k = \hat{p}_k(1-\hat{p}_k)/n_k$.  Find the $\hat{V}_k$, and then re-do the estimation in (4a) and (4b) where the squared error for $p_k$ is divided by $\hat{V}_k$.  How much do the parameter estimates change?  How much do the plotted curves in (4c) change?
```{r}
idx <- !(is.na(p.vec) | p.vec == 0)
wt <- p.vec*(1-p.vec)/k.vec
m.w1 <- lm(p ~ k, data = df.p[idx,], weight = 1/wt[idx])
summary(m.w1)

m.w2 <- nls(p ~ f(k, a, b), data = df.p[idx,], 
            start = list(a = 0, b = -0.2), weight = 1/wt[idx])
summary(m.w2)
```

```{r}
w1 = predict(m.w1, newdata=data.frame(k=c(0:20)))
w2 = predict(m.w2, newdata=data.frame(k=c(0:20)))

df.w <- data.frame(k = 0:20, p = p.vec)
df.w <- mutate(df.w, linear = w1, logistic = w2)
dft <- df.w  |>  gather(model, res, -k)  |>  na.omit()
dft |>  ggplot() + geom_point(aes(x = k, y = res, color = model), size = 3) +
      geom_line(aes(x = k, y = res, color = model), linewidth = 1) +
      labs(y = "Probability",
           title = "Prediction of linear and logistic models with weight")

```

** The change of plots**: Both curves are more fitting when $k$ is small and seem to somehow neglect the dots in the middle-top. From my analysis, the reason is that smaller samples cause larger estimated variances, which reduce the weight of these points. And there are more observations when $k$ is small, which adds to the weight.