---
title: "Homework 1"
author: "Zhang Zile,3230104237"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r setup, message = F, include=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_knit$set(eval.after = 'fig.cap')
library(tidyverse)
library(DAAG)
library(ggpubr)
```

1. The Iowa data set `iowa.csv` is a toy example that summarises the yield of wheat (bushels per acre) for the state of Iowa between 1930-1962. In addition to yield, year, rainfall and temperature were recorded as the main predictors of yield.
    a. First, we need to load the data set into R using the command `read.csv()`. Use the help function to learn what arguments this function takes. Once you have the necessary input, load the data set into R and make it a data frame called `iowa.df`.
    b. How many rows and columns does `iowa.df` have? 
    c. What are the names of the columns of `iowa.df`?
    d. What is the value of row 5, column 7 of `iowa.df`?
    e. Display the second row of `iowa.df` in its entirety.
```{r}
# a
iowa.df<-read.csv("data/Iowa.csv", sep = ';', header=T)
# b
nrow(iowa.df)  #33
ncol(iowa.df)  #10
# c
colnames(iowa.df) #"Year"  "Rain0" "Temp1" "Rain1" "Temp2" "Rain2" "Temp3" "Rain3" "Temp4" "Yield"
# d
iowa.df[5,7] #79.7
# e
iowa.df[,2]
```

2. Syntax and class-typing.
    a. For each of the following commands, either explain why they should be errors, or explain the non-erroneous result.

```
vector1 <- c("5", "12", "7", "32")
```
* Correct,it means vector1 is assigned by a vector which contains 4 characters.
The result is:
```
> vector1 
[1] "5"  "12" "7"  "32"
```
```
max(vector1)
```
* Correct,it returns the maximum element in the vector.
The result is:
```
> max(vector1)
[1] "7"
```

```
sort(vector1)
```
Correct,it sorts the vector in order.The result is:
```
> sort(vector1)
[1] "12" "32" "5"  "7" 
```
```
sum(vector1)
```
* Error,because the type of elements are character and the arguments of sum should be numeric or complex or logical vectors.
    b. For the next series of commands, either explain their results, or why they should produce errors.

```
vector2 <- c("5",7,12)
vector2[2] + vector2[3]

dataframe3 <- data.frame(z1="5",z2=7,z3=12)
dataframe3[1,2] + dataframe3[1,3]

list4 <- list(z1="6", z2=42, z3="49", z4=126)
list4[[2]]+list4[[4]]
list4[2]+list4[4]
```
* Error.A vector can contain elements of only one type.When vector2 is assigned in this example,the entire vector gets converted to character.As a result,you can not do  "vector2[2] + vector2[3]" because their type are characters.

* Correct.It creates a dataframe, the elements are of data types character, numeric, numeric separately  and numeric variable can be added.The result is:
```
> dataframe3[1,2] + dataframe3[1,3]
[1] 19
```
* "list4[[2]]+list4[[4]]" is correct and the result is 168 because numeric variable can be added.But the type of "list4[2]" and "list4[4]" is still list so they can not be added.

3. Working with functions and operators.
    a. The colon operator will create a sequence of integers in order. It is a special case of the function `seq()` which you saw earlier in this assignment. Using the help command `?seq` to learn about the function, design an expression that will give you the sequence of numbers from 1 to 10000 in increments of 372. Design another that will give you a sequence between 1 and 10000 that is exactly 50 numbers in length.
    b. The function `rep()` repeats a vector some number of times. Explain the difference between `rep(1:3, times=3)` and `rep(1:3, each=3)`.
```{r}
# a
  seq(1,10000,372) 
  seq(1,10000,length.out=50)
# b
  rep(1:3,times=3)
  rep(1:3,each=3)
```
* When comes to "rep(1:3,times=3)",the times argument specifies how many times the entire vector should be repeated.
* However,the each argument specifies how many times each individual element of the vector should be repeated.

MB.Ch1.2. The orings data frame gives data on the damage that had occurred in US space shuttle launches prior to the disastrous Challenger launch of 28 January 1986. The observations in rows 1, 2, 4, 11, 13, and 18 were included in the pre-launch charts used in deciding whether to proceed with the launch, while remaining rows were omitted.

Create a new data frame by extracting these rows from orings, and plot total incidents against temperature for this new data frame. Obtain a similar plot for the full data set.

```{r}
data(orings)
prelaunch <- orings[c(1,2,4,11,13,18),]
plot(prelaunch$Temperature, prelaunch$Total) # new dataframe
plot(orings$Temperature, orings$Total) #full data set
```
MB.Ch1.4. For the data frame `ais` (DAAG package)

(a) Use the function `str()` to get information on each of the columns. Determine whether any of the columns hold missing values.
```{r}
data(ais)
str(ais)
complete.cases(t(ais)) #columns
```
(b) Make a table that shows the numbers of males and females for each different sport. In which sports is there a large imbalance (e.g., by a factor of more than 2:1) in the numbers of the two sexes?
```{r}
ais %>% group_by(sport) %>% 
  summarise(sex_f = sum(sex == "f"),sex_m = sum(sex == "m")) %>% 
  mutate(ratio= sex_f / sex_m) %>% 
  mutate(isbalance=(ratio > 0.5) & (ratio < 2))
```
* Gym,Netball,T_Sprnt and W_Polo are not balance.

MB.Ch1.6.Create a data frame called Manitoba.lakes that contains the lake's elevation (in meters above sea level) and area (in square kilometers) as listed below. Assign the names of the lakes using the `row.names()` function.

```{=latex}
\begin{table}[h!]
  \begin{center}
\begin{tabular}{l|r|r} \hline
                & elevation & area \\ \hline
Winnipeg        &     217   & 24387 \\
Winnipegosis    &     254   & 5374 \\
Manitoba        &     248   & 4624 \\
SouthernIndian  &     254   & 2247 \\
Cedar           &     253   & 1353 \\
Island          &     227   & 1223 \\
Gods            &     178   & 1151 \\
Cross           &     207   & 755 \\
Playgreen       &    217    & 657 \\\hline
\end{tabular}
\end{center}
\end{table}
```
```{r}
Manitoba.lakes <- data.frame(elevation = c(217, 254, 248, 254, 253, 227, 178, 207, 217),
                             area = c(24387, 5374, 4624, 2247, 1353, 1223, 1151, 755, 657),
                             row.names = c("Winnipeg","Winnipegosis","Manitoba",
                                           "SouthernIndian","Cedar","Island",
                                           "Gods","Cross","Playgreen"))

```

(a) Use the following code to plot `log2(area)` versus elevation, adding labeling information (there is an extreme value of area that makes a logarithmic scale pretty much essential):

```{r,fig.width=7,fig.height=6,fig.cap="\\label{fig:figs}Use log2(area) to respond to skewness towards large values.The y-axis shows the base-2 logarithm of lake area, so that each unit increase corresponds to a doubling in actual surface area. Labels to the left of each point indicate the lake's surface area in square kilometers; labels to the right show the lake's name."}
attach(Manitoba.lakes)
plot(log2(area) ~ elevation, pch=16, xlim=c(170,280))
# NB: Doubling the area increases log2(area) by 1.0
text(log2(area) ~ elevation, labels=row.names(Manitoba.lakes), pos=4)
text(log2(area) ~ elevation, labels=area, pos=2) 
title("Manitoba's Largest Lakes")
```

Devise captions that explain the labeling on the points and on the y-axis. It will be necessary to explain how distances on the scale relate to changes in area.

(b) Repeat the plot and associated labeling, now plotting area versus elevation, but specifying `ylog=TRUE` in order to obtain a logarithmic y-scale.
```{r,fig.cap="\\label{fig:figs}The y-axis now displays real area values but uses a logarithmic scale to accommodate the large range of sizes. As before, labels to the left indicate surface area, and labels to the right show lake names. This plot is more intuitive than the log-transformed axis but retains the advantage of visualizing both small and large lakes effectively."}
plot(area ~ elevation, pch=16, xlim=c(170,280), ylog=T)
text(area ~ elevation, labels=row.names(Manitoba.lakes), pos=4, ylog=T)
text(area ~ elevation, labels=area, pos=2, ylog=T) 
title("Manitoba's Largest Lakes")

```

MB.Ch1.7. Look up the help page for the R function `dotchart()`. Use this function to display the areas of the Manitoba lakes (a) on a linear scale, and (b) on a logarithmic scale. Add, in each case, suitable labeling information.
```{r}
dotchart(area, labels = rownames(Manitoba.lakes), 
         main = "Manitoba’s Largest Lakes",
         xlab = "area")
dotchart(log2(area), labels = rownames(Manitoba.lakes), 
         main = "Manitoba’s Largest Lakes",
         xlab = "log2(area)")
```

MB.Ch1.8. Using the `sum()` function, obtain a lower bound for the area of Manitoba covered by water.
```{r}
sum(area)
```