---
title: "Homework 12"
author: "张子乐, 3230104237"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage[fontset=fandol]{ctex}
  - \usepackage{float}
---

```{r setup, include=FALSE}
Sys.setenv(FONTCONFIG_PATH = "C:/Users/admin/AppData/Roaming/TinyTeX/tlpkg/tlpostcode/xetex/conf")
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  dev = "cairo_pdf",
  comment = "",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)

library(ggplot2)
library(gridExtra)
```

# 1. Background

The MNIST data set contains grayscale images of handwritten digits. Each image has size
$28 \times 28$, so one image can be represented as a vector with
$D = 784$ pixel values. 

In this homework, the grayscale values are converted into **binary**
values: pixels larger than a fixed threshold are set to 1, and the remaining pixels are
set to 0.

First, we use the digits 3, 4, and 5 only. Although their labels are used to select the data,
the mixture model itself is fitted without using the labels. 

The initial goal is to estimate $K = 3$ latent Bernoulli components by the expectation-maximization (EM) algorithm and
then reshape the estimated Bernoulli parameters back to images. 

# 2. Model

Let $\mathbf{x}_n \in \{0,1\}^D$ be the binarized and flattened image for observation
$n$. A Bernoulli mixture model assumes that each observation is generated from one of
$K$ latent components:

$$
p(\mathbf{x}_n)=\sum_{k=1}^K \pi_k p(\mathbf{x}_n|\boldsymbol{\mu}_k),
$$

where $\pi_k$ is the mixing proportion of component $k$, $\sum_{k=1}^K \pi_k = 1$, and
$\boldsymbol{\mu}_k = (\mu_{k1}, \ldots, \mu_{kD})$ contains the Bernoulli probabilities
for the pixels in component $k$. The component likelihood is

$$
p(\mathbf{x}_n|\boldsymbol{\mu}_k)
=
\prod_{j=1}^{D}
\mu_{kj}^{x_{nj}}(1-\mu_{kj})^{1-x_{nj}}.
$$

The log-likelihood is

$$
\log p(X)
=
\sum_{n=1}^{N}
\log
\left[
\sum_{k=1}^{K}
\pi_k p(\mathbf{x}_n|\boldsymbol{\mu}_k)
\right].
$$

# 3. Data Preparation and Helper Functions

## 3.1 Hyperparameters

- In the experiment below, the digit images are first converted into binary vectors by
setting all pixels whose grayscale values exceed `127` to 1 and setting the remaining
pixels to 0. 

- We fit a data set containing the handwritten digits `3`, `4`, and `5`,
with at most 600 images sampled from each digit, using a mixture of `K = 3` Bernoulli
distributions. 

- The EM algorithm is run for at most 100 iterations.

- The mixing coefficients are initialized as $\pi_k = 1/K$, and the Bernoulli
parameters $\mu_{kj}$ are initialized independently from the uniform distribution
$U(0.25, 0.75)$.

```{r helper-functions}
selected_digits <- c(3, 4, 5) # 选择3,4,5三个数字
K <- length(selected_digits)
max_per_digit <- 600 # 每个数字最多选取600个样本
threshold <- 127 # 灰度值二值化的阈值
max_iter <- 100 # EM算法的最大迭代次数
tol <- 1e-6 # 收敛阈值
eps <- 1e-6 
set.seed(42)
```

## 3.2 Helper Functions

```{r}
# 数值稳定性
log_sum_exp <- function(log_values) {
  max_values <- apply(log_values, 1, max)
  max_values + log(rowSums(exp(log_values - max_values)))
}

clip_prob <- function(x, eps = 1e-6) {
  pmin(pmax(x, eps), 1 - eps)
}

image_to_df <- function(values) {
  image_matrix <- matrix(values, nrow = 28, ncol = 28, byrow = TRUE)
  df <- as.data.frame(as.table(image_matrix))
  names(df) <- c("row", "col", "value")
  df$row <- as.integer(df$row)
  df$col <- as.integer(df$col)
  df$x <- df$col
  df$y <- 29 - df$row
  df
}

plot_digit_image <- function(values, title) {
  ggplot(image_to_df(values), aes(x = x, y = y, fill = value)) +
    geom_raster() +
    coord_fixed() +
    scale_fill_gradient(low = "white", high = "black", limits = c(0, 1)) +
    labs(title = title) +
    theme_void(base_size = 11) +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
}
```


## 3.3 Data Loading and Reading

```{r}
# 下载MNIST数据集
mnist_dir <- file.path("data", "mnist")
download_mnist <- function(dest_dir = mnist_dir) {
  dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
  files <- c(
    images = "train-images-idx3-ubyte.gz",
    labels = "train-labels-idx1-ubyte.gz"
  )
  base_url <- "https://storage.googleapis.com/cvdf-datasets/mnist"
  for (file_name in files) {
    dest <- file.path(dest_dir, file_name)
    if (!file.exists(dest)) {
      url <- paste(base_url, file_name, sep = "/")
      tryCatch(
        download.file(url, destfile = dest, mode = "wb", quiet = TRUE),
        error = function(e) {
          stop(
            "Could not download MNIST automatically. ",
            "Please place train-images-idx3-ubyte.gz and ",
            "train-labels-idx1-ubyte.gz in ",
            normalizePath(dest_dir, winslash = "/", mustWork = FALSE),
            ". Original error: ",
            conditionMessage(e),
            call. = FALSE
          )
        }
      )
    }
  }
  paths <- file.path(dest_dir, unname(files))
  names(paths) <- names(files)
  invisible(paths)
}

# 读取MNIST图像和标签
read_mnist_images <- function(path) {
  con <- gzfile(path, "rb")
  on.exit(close(con))
  magic <- readBin(con, integer(), n = 1, size = 4, endian = "big")
  n_images <- readBin(con, integer(), n = 1, size = 4, endian = "big")
  n_rows <- readBin(con, integer(), n = 1, size = 4, endian = "big")
  n_cols <- readBin(con, integer(), n = 1, size = 4, endian = "big")
  if (magic != 2051 || n_rows != 28 || n_cols != 28) {
    stop("Unexpected MNIST image file format.", call. = FALSE)
  }
  pixels <- readBin(
    con,
    what = integer(),
    n = n_images * n_rows * n_cols,
    size = 1,
    signed = FALSE
  )
  matrix(pixels, nrow = n_images, byrow = TRUE)
}

read_mnist_labels <- function(path) {
  con <- gzfile(path, "rb")
  on.exit(close(con))
  magic <- readBin(con, integer(), n = 1, size = 4, endian = "big")
  n_labels <- readBin(con, integer(), n = 1, size = 4, endian = "big")
  if (magic != 2049) {
    stop("Unexpected MNIST label file format.", call. = FALSE)
  }
  readBin(con, what = integer(), n = n_labels, size = 1, signed = FALSE)
}

mnist_files <- download_mnist()
images <- read_mnist_images(mnist_files["images"])
labels <- read_mnist_labels(mnist_files["labels"])

selected_index <- unlist(lapply(selected_digits, function(digit) {
  candidates <- which(labels == digit)
  sample(candidates, min(max_per_digit, length(candidates)))
}))

selected_images <- images[selected_index, , drop = FALSE]
selected_labels <- labels[selected_index]
binary_images <- 1L * (selected_images > threshold)
```

The following figure shows one binarized example from each selected digit.

```{r sample-images, fig.height=2.4}
sample_plots <- lapply(selected_digits, function(digit) {
  first_index <- which(selected_labels == digit)[1]
  plot_digit_image(binary_images[first_index, ], paste("Digit", digit))
})
do.call(grid.arrange, c(sample_plots, nrow = 1))
```


## 3.4 EM Algorithm Implementation

The latent variable $z_{nk}$ indicates whether observation $n$ belongs to component
$k$. EM alternates between computing the posterior **responsibilities** and **updating the parameters**.

## 3.4.1 E-step

In the E-step, compute the resposibilities

$$
\gamma_{nk}
=
\frac{\pi_k p(\mathbf{x}_n|\boldsymbol{\mu}_k)}
{\sum_{\ell=1}^K \pi_\ell p(\mathbf{x}_n|\boldsymbol{\mu}_\ell)}
$$

## 3.4.2 M-step

In the M-step, first define the effective sample size

$$
N_k=\sum_{n=1}^N \gamma_{nk}
$$

Then update

$$
\pi_k=\frac{N_k}{N},
\qquad
\mu_{kj}
=
\frac{1}{N_k}
\sum_{n=1}^N
\gamma_{nk}x_{nj}
$$

The computation below uses log probabilities for numerical stability.

```{r}
EM <- function(x, K, max_iter = 100, tol = 1e-6, eps = 1e-6) {
  n <- nrow(x) # n个样本
  d <- ncol(x) # d维
  # Initialize pi and mu
  pi <- rep(1 / K, K)
  mu <- matrix(runif(K * d, min = 0.25, max = 0.75), nrow = K, ncol = d)
  log_likelihood <- numeric()
  responsibilities <- matrix(NA_real_, nrow = n, ncol = K)

  for (iter in seq_len(max_iter)) {
    mu <- clip_prob(mu, eps)
    pi <- clip_prob(pi, eps)
    pi <- pi / sum(pi)

    # E-step
    log_component <- x %*% t(log(mu)) + (1 - x) %*% t(log(1 - mu)) # log p(x_n | mu_k)
    log_joint <- sweep(log_component, 2, log(pi), "+") # 按列加上 log pi_k
    log_norm <- log_sum_exp(log_joint)
    responsibilities <- exp(log_joint - log_norm) # responsibilities

    current_ll <- sum(log_norm) # 当前的 log-likelihood
    log_likelihood <- c(log_likelihood, current_ll)

    # M-step
    nk <- colSums(responsibilities) # effective number of data points 
    pi <- nk / n
    mu <- t(responsibilities) %*% x # K x D
    mu <- sweep(mu, 1, nk, "/")
    mu <- matrix(mu, nrow = K, ncol = d)

    # 判断是否收敛
    if (iter > 1) {
      improvement <- abs(log_likelihood[iter] - log_likelihood[iter - 1])
      scale <- abs(log_likelihood[iter - 1]) + eps
      if (improvement / scale < tol) {
        break
      }
    }
  }

  mu <- clip_prob(mu, eps)
  pi <- clip_prob(pi, eps)
  pi <- pi / sum(pi)
  log_component <- x %*% t(log(mu)) +
    (1 - x) %*% t(log(1 - mu))
  log_joint <- sweep(log_component, 2, log(pi), "+")
  log_norm <- log_sum_exp(log_joint)
  responsibilities <- exp(log_joint - log_norm)
  final_log_likelihood <- sum(log_norm)

  list(
    pi = pi,
    mu = clip_prob(mu, eps),
    responsibilities = responsibilities,
    log_likelihood = log_likelihood,
    final_log_likelihood = final_log_likelihood,
    iterations = length(log_likelihood)
  )
}
```

# 4. Experiment Results

```{r fit-em}
em_fit <- EM(
  x = binary_images,
  K = K,
  max_iter = max_iter,
  tol = tol,
  eps = eps
)
cat("Weights (pi):", round(em_fit$pi, 4), "\n")
```

## 4.1 Convergence

The log-likelihood curve is used to check convergence.

```{r likelihood-plot}
likelihood_df <- data.frame(
  iteration = seq_along(c(em_fit$log_likelihood, em_fit$final_log_likelihood)),
  log_likelihood = c(em_fit$log_likelihood, em_fit$final_log_likelihood)
)

ggplot(likelihood_df, aes(x = iteration, y = log_likelihood)) +
  geom_line(linewidth = 0.7, color = "#000000") +
  geom_point(size = 1.5, color = "#000000") +
  labs(
    x = "Iteration",
    y = "Log-likelihood"
  ) +
  theme_minimal(base_size = 12)
```

## 4.2 Reconstructed Digit Templates

Each row of $\boldsymbol{\mu}$ is a vector of 784 Bernoulli probabilities. After
reshaping it into a $28 \times 28$ matrix, dark pixels indicate locations that are
estimated to be active with high probability under that component.We can distinguish the components 
by visually comparing the reconstructed templates with the shapes of the selected handwritten digits.

```{r estimated-templates, fig.height=2.4}
template_plots <- lapply(seq_len(K), function(k) {
  title <- paste0("Component ", k, ", pi = ", round(em_fit$pi[k], 3))
  plot_digit_image(em_fit$mu[k, ], title)
})
do.call(grid.arrange, c(template_plots, nrow = 1))
```

## 4.3 Comment on Results

- The reconstructed templates suggest that Component 1 approximately corresponds to digit 5, Component 2 to digit 4, 
and Component 3 to digit 3. 

- The ordering of these components is arbitrary because the EM algorithm is fitted without using the digit labels. Although the sampled data contain approximately equal numbers of digits 3, 4, and 5, 
the estimated mixture weights do not have to be exactly 1/3, because the latent components are inferred from image patterns rather than fixed by the observed labels.

## 4.4 Comparison with a Single Bernoulli Distribution

As a comparison, we also fit the same data set using a single multivariate Bernoulli
distribution, again using maximum likelihood. This amounts to simply averaging the
counts in each pixel, which produces a blurred average template, whereas the mixture model separates the major digit patterns into interpretable components.

```{r single-bernoulli-template, fig.height=2.4}
single_mu <- colMeans(binary_images)
plot_digit_image(single_mu, "Single Bernoulli")
```

# 5. Choosing the Number of Components

To investigate whether the number of latent components can be inferred without
prespecifying $K$, we fitted Bernoulli mixture models with several candidate values
of $K$ and compared their AIC/BIC values.

For a Bernoulli mixture model with $K$ components and $D$ pixels, the number of free
parameters is

$$
p = K D + (K - 1),
$$

where $K D$ counts the Bernoulli probabilities and $K - 1$ counts the free mixing
coefficients. Therefore,

$$
\mathrm{AIC} = 2p - 2\ell,
\qquad
\mathrm{BIC} = \log(N)p - 2\ell,
$$

where $\ell$ is the maximized log-likelihood.

Each candidate value of $K$ is fitted multiple times and the run with the
largest final log-likelihood is retained and used to compute AIC/BIC for each $K$.

```{r aic-bic-comparison}
set.seed(4237)
candidate_K <- 1:10 # 评估K=1到10的模型
n_starts <- 3 # 重复运行3次EM算法，选择对数似然最大的结果
n <- nrow(binary_images)
d <- ncol(binary_images)
best_models <- vector("list", length(candidate_K))
names(best_models) <- as.character(candidate_K)
model_selection_rows <- vector("list", length(candidate_K))

for (i in seq_along(candidate_K)) {
  k <- candidate_K[i]
  fits <- lapply(seq_len(n_starts), function(start) {
    EM(
      x = binary_images,
      K = k,
      max_iter = max_iter,
      tol = tol,
      eps = eps
    )
  })
  log_likelihoods <- vapply(
    fits,
    function(fit) fit$final_log_likelihood,
    numeric(1)
  )
  best_start <- which.max(log_likelihoods)
  best_fit <- fits[[best_start]]
  best_models[[as.character(k)]] <- best_fit

  n_params <- k * d + (k - 1)
  log_likelihood <- best_fit$final_log_likelihood
  model_selection_rows[[i]] <- data.frame(
    K = k,
    best_start = best_start,
    log_likelihood = log_likelihood,
    parameters = n_params,
    AIC = 2 * n_params - 2 * log_likelihood,
    BIC = log(n) * n_params - 2 * log_likelihood,
    iterations = best_fit$iterations
  )
}

model_selection <- do.call(rbind, model_selection_rows)
# 不同K下的对数似然/AIC/BIC
model_selection
```

```{r aic-bic-plot, fig.height=3.2}
selection_plot_df <- rbind(
  data.frame(K = model_selection$K, criterion = "AIC", value = model_selection$AIC),
  data.frame(K = model_selection$K, criterion = "BIC", value = model_selection$BIC)
)
ggplot(selection_plot_df, aes(x = K, y = value, color = criterion)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  scale_x_continuous(breaks = candidate_K) +
  labs(
    x = "K",
    y = "Value",
    color = NULL
  ) +
  theme_minimal(base_size = 12)
```

```{r}
# 选择使AIC/BIC最小的K
k_AIC <- model_selection$K[which.min(model_selection$AIC)]
k_BIC <- model_selection$K[which.min(model_selection$BIC)]
cat("Selected K by AIC:", k_AIC, "\n")
cat("Selected K by BIC:", k_BIC, "\n")
```

```{r}
em_AIC <- best_models[[as.character(k_AIC)]]
em_BIC <- best_models[[as.character(k_BIC)]]
cat("Weights (pi) for AIC:", round(em_AIC$pi, 4), "\n")
cat("Weights (pi) for BIC:", round(em_BIC$pi, 4), "\n")
```

```{r best-k7-templates, fig.height=4.8}
template_plots_BIC <- lapply(seq_len(k_BIC), function(k) {
  title <- paste0("pi = ", round(em_BIC$pi[k], 3))
  plot_digit_image(em_BIC$mu[k, ], title)
})
do.call(grid.arrange, c(template_plots_BIC, nrow = 2))
```

## Conclusion

- The AIC reaches its minimum at $K = `r k_AIC `$, whereas the BIC reaches its minimum at $K = `r k_BIC `$. Each value of $K$ is fitted with `r n_starts ` random starts, and the AIC/BIC values are computed from the run with the largest log-likelihood.

- Although the data contain only three digit labels, the number of mixture components does not need to equal three. Different handwriting styles within the same digit may be captured by separate components.

- The saved best model for $K = 7$ is plotted above. Some components represent different styles of the same digit, such as upright and slanted forms.

- Therefore, the additional components should be interpreted as handwriting-style subtypes rather than new digit classes.
